---
title: "Comparison of comparisons"
author: "Suzie Brown"
date: "22 November 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(BradleyTerry2)
library(StatRank)
library(DescTools)
```

## Define functions
```{r}
# to simulate ratings-type data
make.ratings <- function(n.alts, n.raters, strengths=NULL, sd.strengths=1, sd.noise=0.1) {
  if (!is.null(strengths) && length(strengths)!=n.alts){stop("length of strengths vector is not equal to number of alternatives")}
  if (is.null(strengths)){strengths <- rnorm(n.alts, 0, sd.strengths)}
  ratings <- replicate(n.raters, strengths + rnorm(n.alts, 0, sd.noise))
  list(ratings=ratings, strengths=strengths)
}

# to convert ratings to rankings
ratings2ranking <- function(ratings) {
  invrank <- apply(ratings, 2, order, decreasing=TRUE)
  rank <- apply(invrank, 2, order)
  rank
}

# to convert rankings to pairwise summary
ranking2pairs <- function(ranking) {
  n.alts <- nrow(ranking)
  w <- matrix(NA, nrow=n.alts, ncol=n.alts)
  for (i in 1:n.alts) {
    for (j in 1:n.alts) {
      w[i,j] <- sum(ranking[i,]<ranking[j,])
    }
  }
  w
}

# to normalise the strength vectors
normalise <- function(x, k=2) {
  xcentre <- x - mean(x)
  xcentre/sum(xcentre^k)
}
```

# Preliminary experiments

## Simulate some data and convert to each data type
```{r}
n.alts <- 20
foo <- make.ratings(n.alts=n.alts, n.raters=10, strengths=1:n.alts, sd.noise=5)
strengths <- foo$strengths
true.ranks <- order(order(strengths, decreasing=TRUE))

ratings <- foo$ratings
ranks <- ratings2ranking(ratings)
pairs <- ranking2pairs(ranks)
rownames(pairs) <- 1:nrow(pairs)
colnames(pairs) <- 1:nrow(pairs)
```

## Fit each model to the corresponding data type

### Ratings-type data (linear model)
```{r}
s.hat.ratings <- rowMeans(ratings) # estimated strengths (MLE)
r.hat.ratings <- order(order(s.hat.ratings, decreasing=TRUE))# estimated rankings
```

### Rankings-type data (Plackett-Luce model)
```{r}
rev.ranks <- apply(ranks,2,order, decreasing=TRUE)
PLfit <- Estimation.PL.MLE(t(rev.ranks))
s.hat.ranks <- log(PLfit$Mean) # we need to log the output from PL
r.hat.ranks <- order(order(PLfit$Mean, decreasing=TRUE))
```

### Pair comparison-type data (Bradley-Terry model)
```{r}
pairs.long <- countsToBinomial(pairs) # change contingency table to tall data format
BTfit <- BTm(outcome=as.matrix(pairs.long[,3:4]), player1=pairs.long[,1], player2=pairs.long[,2])
s.hat.pairs <- c(0,BTfit$coefficients)
r.hat.pairs <- order(order(c(1,as.vector(BTfit$coefficients)), decreasing=T))
```

## Compare the (normalised) estimated strengths
Some information has been lost in converting data types, so it is not possible to recover the absolute strengths. We can recover relative strengths. To compare the output of each method we will centre and scale the strength vectors.
```{r}
S.true <- scale(strengths)
S.rating <- scale(s.hat.ratings)
S.pair <- scale(s.hat.pairs)
S.rank <- scale(s.hat.ranks)
```
Now make a plot.
```{r}
plot(S.true, type='l', lwd=2, col="grey", ylim=range(S.true)+0.5*c(-1,1), ylab="Strength", main="Estimated strength of each alternative")
points(S.rating, type='l', lwd=2, col=2)
points(S.rank, type='l', lwd=2, col=4)
points(S.pair, type='l', lwd=2, col="purple")
legend("topleft", c("true", "ratings: LM", "ranks: PL", "pairs: BT"), lwd=2, col=c('grey',2,4,'purple'), lty=1)
```

## Performance Metrics
Kendall's tau distance is the number of discordant pairs between two vectors; we will use this as a distance measure (metric?) for rankings.
```{r}
rankdist.pairs <- ConDisPairs(table(true.ranks, r.hat.pairs))$D
rankdist.ranks <- ConDisPairs(table(true.ranks, r.hat.ranks))$D
rankdist.ratings <- ConDisPairs(table(true.ranks, r.hat.ratings))$D

MSE.pairs <- mean((strengths-s.hat.pairs)^2)
MSE.ranks <- mean((strengths-s.hat.ranks)^2)
MSE.ratings <- mean((strengths-s.hat.ratings)^2)

bias.pairs <- mean(strengths-s.hat.pairs)
bias.ranks <- mean(strengths-s.hat.ranks)
bias.ratings <- mean(strengths-s.hat.ratings)

perftable <- data.frame(rankdist=c(rankdist.ratings,rankdist.ranks, rankdist.pairs), strengthMSE=c(MSE.ratings,MSE.ranks,MSE.pairs), strengthbias=c(bias.ratings,bias.ranks, bias.pairs))
rownames(perftable) <- c("Ratings: LM", "Ranks: PL", "Pairs: BT")

kable(perftable, digits=3, row.names=TRUE, col.names=c("Kendall tau", "MSE strengths", "Bias strengths"))